Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2015-12-15 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Rewrite titles of sections
Closes #65
Ditto fig 7 refs #70

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of blue indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in red). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The red dot and line show the average and standard
  deviation, the shade of blue the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
\end{figure}

>>>>

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of blue (or gray in grayscale) indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade of gray). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the shades of gray the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}
